---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 4 - Due date 02/17/22"
author: "Abbhijith Hari Gopal"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
  You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp21.Rmd"). Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(ggplot2)
library(forecast)
library(Kendall)
library(readxl)
library(tseries) 
library(lubridate)
library(dplyr)
```

## Questions 

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production". 


```{r}
#Importing data set - using xlsx package
getwd()
data = read_excel("./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", skip = 12, sheet="Monthly Data",col_names=FALSE)
data_of_interest = data[, 5]
date_of_interest = data[, 1]
head(data_of_interest, 15) #As a check
head(date_of_interest, 15) #As a check

colnames(data_of_interest) = c("Total Renewable Energy Production")
colnames(date_of_interest) = c("Date")
data_of_interest$`Total Renewable Energy Production` = as.numeric(data_of_interest$`Total Renewable Energy Production`)
class(data_of_interest) #As a check

new_data = cbind(date_of_interest, data_of_interest)
head(new_data, 10) #As a check

ncolumns = ncol(new_data)
nmonths = nrow(new_data)

ts_data = ts(new_data[, 2], start = c(1973, 1), frequency = 12)
head(ts_data, 50) #As a check
class(ts_data) #As a check

```


## Stochastic Trend and Stationarity Tests

### Q1

Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments:
* *x* vector containing values to be differenced;
* *lag* integer indicating with lag to use;
* *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series Do the series still seem to have trend?

```{r original and differenced series plot}
plot(new_data[,2], type="l",col="blue",ylab="Trillion Btu", main = "Original Series plot")
differenced_series = diff(new_data[, 2], lag = 1, differences = 1)
plot(differenced_series, type="l",col="red",ylab="Trillion Btu", main = "Differenced Series plot")
head(differenced_series, 10) #As check
```
**From the plots, the original series seems to have a positive, increasing trend. The Differenced series, on the other hand, doesn't seem to have any trend.**
### Q2

Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in A3 using linear regression. (Hint: Just copy and paste part of your code for A3)

Copy and paste part of your code for A3 where you compute regression for Total Energy Production and the detrended Total Energy Production

```{r LR_detrended}
t = c(1:nmonths)

renewable_linear_trend = lm(new_data$`Total Renewable Energy Production`~t)
beta0 = as.numeric(renewable_linear_trend$coefficients[1])
beta1 = as.numeric(renewable_linear_trend$coefficients[2])

renewable_detrend = new_data$`Total Renewable Energy Production` - (beta0 + beta1*t)

linear_trend_model=lm(renewable_detrend~t) 
linear_trend_beta0 = as.numeric(linear_trend_model$coefficients[1])
linear_trend_beta1 = as.numeric(linear_trend_model$coefficients[2])
summary(linear_trend_model) #detrended data

print(paste("The slope is", linear_trend_beta1, "and intercept is", linear_trend_beta0))
```

```{r LR_differenced}
n2 = nrow(new_data) - 1
t1 = c(1:n2)
diff_lin_trend = lm(differenced_series~t1)
beta0 = as.numeric(diff_lin_trend$coefficients[1])
beta1 = as.numeric(diff_lin_trend$coefficients[2])
summary(diff_lin_trend) #differenced data

print(paste("The slope is", beta1, "and intercept is", beta0))

```
From the output of the regression analysis for detrended and differenced series, we can observe that for detrended series, the slope is `r linear_trend_beta1` and intercept is `r linear_trend_beta0`; for differenced series, the slope is `r beta1` and intercept is `r beta0`. \
Interpretation: For detrended series by regression: The renewable energy production increases by `r linear_trend_beta1` trillion Btu in 1 year, keeping all other variables constant. \
For differenced series: The renewable energy production increases by `r beta1` trillion Btu in 1 year, keeping all other variables constant.

### Q3
Create a data frame with 4 columns: month, original series, detrended by Regression Series and differenced series. Make sure you properly name all columns. Also note that the differenced series will have only 584 rows because you loose the first observation when differencing. Therefore, you need to remove the first observations for the original series and the detrended by regression series to build the new data frame.

```{r}
#Data frame - remember to note include January 1973
renewable_detrend_df = data.frame(renewable_detrend)
nrow(renewable_detrend_df) #As check
differenced_series_df = data.frame(differenced_series)
nrow(differenced_series_df) #As check
new_data = data.frame(new_data)
nrow(new_data) #As check

colnames(renewable_detrend_df) = c("Detrended data")
colnames(differenced_series_df) = c("Differenced data")
truncated_new_data = new_data[-1,]
truncated_renewable_detrend = renewable_detrend_df[-1,]
new_df = cbind(truncated_new_data, truncated_renewable_detrend, differenced_series_df) #new data frame with all columns of interest merged. 
head(new_df, 5) #As check
class(new_df) #As check
nrow(new_df) #As check
```

### Q4
Using ggplot() create a line plot that shows the three series together. Make sure you add a legend to the plot.
```{r}
ggplot(data = new_df) + 
  ylab("Trillion BTu") + 
  geom_line(aes(Date, new_df$`Differenced data`, colour = "Differenced data")) +
  geom_line(aes(Date, new_df$Total.Renewable.Energy.Production, colour = "Original Data")) +
  geom_line(aes(Date, new_df$truncated_renewable_detrend, colour = "Detrended data")) +
   scale_colour_manual("", 
                      breaks = c("Differenced data", "Original Data", "Detrended data"),
                      values = c("Differenced data"="green", "Original Data"="red", 
                                 "Detrended data"="blue")) + 
  ggtitle("All 3 series in one plot!")
```


### Q5
Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the Acf() function to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?

```{r}
#Compare ACFs
ts_new_object = ts(new_df[, 2:4], frequency = 12)
head(ts_new_object, 5)
par(mfrow=c(1,3))
Acf(ts_new_object[, 1], lag.max=40,main=paste("Total Renewable Energy Production (original data)"), ylim=c(-0.5,1))
Acf(ts_new_object[, 2], lag.max=40,main=paste("Detrended data (regression)"), ylim=c(-0.5,1))
Acf(ts_new_object[, 3], lag.max=40,main=paste("Differenced Data"), ylim=c(-0.5,1))
  
```
**From the ACF plots, we can observe that differencing seems to remove trend in a more effective way than linear regression. This is because in the detrended ACF plot, detrended by linear regression, it looks like there is still some sort of decreasing trend with increase in lag whereas there isn't any clear trend pattern in the differenced plot**

### Q6
Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both test. Whats the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.

```{r}
SMKtest <- SeasonalMannKendall(ts_data)
print("Results for Seasonal Mann Kendall /n")
print(summary(SMKtest))

print("Results for ADF test /n")
print(adf.test(ts_data, alternative = "stationary"))
```
**From the summary stats for seasonal Mann Kandall test, we can see p-value < 0.05. Therefore, we can reject the null hypothesis which states that there is no deterministic trend in the data. The idea behind the ADF test is to check for existence of stochastic trend and existence unit root in the data set. P value of 0.81 > 0.05 shows that we can't reject the null hypothesis which states that there exists a unit root. Therefore, we need to use a different procedure to remove the trend.**
### Q7

Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. 

```{r}
data_matrix = matrix(ts_data,byrow=FALSE,nrow=12)
data_yearly = colMeans(data_matrix)
head(data_yearly, 50)

my_year = c(year(first(new_data$Date)):year(last(new_data$Date)))
#print(year(last(date_of_interest)))
head(my_year, 50)

data_new_yearly = data.frame(my_year, data_yearly)

ggplot(data_new_yearly, aes(x = my_year, y = data_yearly)) +
  geom_line(color = "red") + 
  geom_smooth(color = "blue", method = "lm")

```

### Q8

Apply the Mann Kendal, Spearman correlation rank test and ADF. Are the results from the test in agreement with the test results for the non-aggregated series, i.e., results for Q6?

```{r}
test1 = ts(data_yearly, start = 1973, frequency = 1)
print("Results of Mann Kendall on average yearly series")
print(summary(MannKendall(test1)))

print("Results from Spearman Correlation")
sp = cor.test(my_year, test1, method="spearman")
print(sp)

print("Results for ADF test on yearly data/n")
print(adf.test(test1, alternative = "stationary"))

```
**After aggregating the data yearly, we can see that for the Seasonal Mann Kandall test, the p value, once again is less than 0.05. Therefore, we have the same conclusion we derived in Q6 - there is a deterministic trend. The Spearmann correlation test suggests that the presence of deterministic test too since its p-value is also less than 0.05. Moreover, the positive value of s-score tells us the trend is an increasing trend. The ADF tests' conclusion is also in line with what we got in Q6. The p-value for the aggregated data by year is 0.4907 > 0.05. Therefore, there exists a unit root in the yearly aggregated data too**
