---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 2 - Due date 01/26/22"
author: "Abbhijith Hari Gopal"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is change "Student Name" on line 4 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp22.Rmd"). Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
library(forecast)#Load/install required package here
library(tseries)
library(dplyr)
library(lubridate)
library(readxl)
library(ggplot2)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the January 2022 Monthly Energy Review. The spreadsheet is ready to be used. Use the command $read.table()$ to import the data in R or $panda.read\_excel()$ in Python (note that you will need to import pandas package). }

```{r}
#Importing data set
data <- read_excel("../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx")
View(data)
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
data_of_interest = data[12:596, 4:6]
colnames(data_of_interest) = c("Total Biomass Energy Production", "Total Renewable Energy Production", "Hydroelectric Power Consumption")
head(data_of_interest)
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
ts_data = ts(data=data_of_interest, start=1, frequency = 12)
ts_data
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
print('Biomass: Mean =')
mean_bio = mean(ts_data[, 1])
mean_bio
print('Total Renewables: Mean =')
mean_rene = mean(ts_data[, 2])
mean_rene
print('Hydro: Mean =')
mean_hydro = mean(ts_data[, 3])
mean_hydro
print('Biomass: Standard deviation =')
std_bio = sd(ts_data[, 1])
std_bio
print('Total Renewables: Standard deviation =')
std_rene = sd(ts_data[, 2])
std_rene
print('Hydro: Standard deviation =')
std_hydro = sd(ts_data[, 3])
std_hydro
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
plot(ts_data[,1],type="l",col="blue",ylab="Trillion Btu",main="Total Biomass Energy Production")
abline(h=mean_bio, col="red")
```

```{r}
plot(ts_data[,2],type="l",col="green",ylab="Trillion Btu",main="Total Renewable Energy Production")
abline(h=mean_rene, col="blue")
```

```{r}
plot(ts_data[,3],type="l",col="red",ylab="Trillion Btu",main="Hydrorelectric Power Consumption")
abline(h=mean_hydro, col="green")
```

## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor(ts_data)
```
*The correlation between Total Biomass Energy Production and Total Renewable Energy Production is a strong positive correlation, the coeffecient being 0.732. The correlation between Hydroelectric Power Consumption and Total Biomass Energy Production is a moderately strong negative correlation, the coefficient being -0.272. The correlation between Hydroelectric Power Consumption and Total Renewable Energy Production is a weak positive correlation, the coefficient being 0.0686.* 

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
acf(x=ts_data[, 1], lag.max = 40)
acf(x=ts_data[, 2], lag.max = 40)
acf(x=ts_data[, 3], lag.max = 40)
```
*The ACF plot for Total Biomass Energy Production shows a decrease with increase in lag. The ACF plot for Total Renewable Energy Production shows a declining trend as well, for the most part, but there seems to be some sort of seasonality in the data, although not clearly observed. The ACF plot for Hydroelectric Power Consumption shows rapid increases and decreases, with increase in lag from 1 to 40, suggesting the unequivocal presence of seasonality.*

## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
pacf(x=ts_data[, 1], lag.max = 40)
pacf(x=ts_data[, 2], lag.max = 40)
pacf(x=ts_data[, 3], lag.max = 40)
```
*The PACF plots basically removes the influence of intermediary correlations, something that the ACF does not. All three PACF plots - Total Biomass Energy Production, Total Renewable Energy Production, and Hydroelectric Power Consumption suggests the presence of seasonality (although minimal for Total Biomass Energy Production & Total Renewable Energy Production). The ACF plots, obtained from Question 6, do not show this unequivocal presence of seasonality. Therefore, if we were to move forward building our ARIMA models by just plotting ACF to check for seasonality, we would be incorrect because, we'll arrive at the conclusion that the models are non-seasonal, meaning stationary - which is clearly wrong. Therefore, PACF plots are crucial to examine the presence of seasonality.*
