---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 3 - Due date 02/08/22"
author: "Abbhijith Hari Gopal"
output: pdf_document
geometry: margin=2.54cm
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Loalibrary(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(readxl)
library(tseries) 
library(lubridate)
```

```{r}
#Importing data set
getwd()
data = read_excel("../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", skip = 12, sheet="Monthly Data",col_names=FALSE)
data_of_interest = data[, 4:6]
date_of_interest = data[, 1]
#date_of_interest = parse_date_time(date_of_interest)

colnames(data_of_interest) = c("Total Biomass Energy Production", "Total Renewable Energy Production", "Hydroelectric Power Consumption")
colnames(date_of_interest) = c("Date")
data_of_interest$`Total Biomass Energy Production` = as.numeric(data_of_interest$`Total Biomass Energy Production`)
data_of_interest$`Total Renewable Energy Production` = as.numeric(data_of_interest$`Total Renewable Energy Production`)
data_of_interest$`Hydroelectric Power Consumption` = as.numeric(data_of_interest$`Hydroelectric Power Consumption`)

new_data = cbind(date_of_interest, data_of_interest[, 1:3])
head(new_data)

ncolumns = ncol(new_data)
nmonths = nrow(new_data)

ts_object = ts(data = new_data[, 2:4], start = c(1973, 1), frequency=12)
ts_object = cbind(date_of_interest, ts_object)
ts_object = ts(ts_object, start = c(1973, 1), frequency=12)

head(ts_object, 15)
class(date_of_interest)
```
 
 
##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)
```{r}
for (i in 2:4)
{
  par(mfrow=c(1,3))
  plot(ts_object[, i], type="l",col="red",ylab="Trillion Btu" ) 
  abline(h=mean(ts_object[, i]), col="blue")
  Acf(ts_object[, i], lag.max=40,main=paste("Column",i,"ACF",sep=" "))
  Pacf(ts_object[, i], lag.max=40,main=paste("Column",i,"PACF",sep=" "))
  
}
```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

**Total Biomass Energy Production and Total Renewable Energy Production seem to have a positive trend whereas, Hydroelectric Power Consumption doesn't appear to have a trend/the trend is not so apparent**

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.
```{r}
t = c(1: nmonths)

linear_trend_model1=lm(ts_object[,2]~t)  
summary(linear_trend_model1)
print(paste("The slope is", linear_trend_model1$coefficients[2], "and intercept is", linear_trend_model1$coefficients[1]))
linear_trend1_beta0 = as.numeric(linear_trend_model1$coefficients[1])
linear_trend1_beta1 = as.numeric(linear_trend_model1$coefficients[2])

linear_trend_model2=lm(ts_object[,3]~t)  
summary(linear_trend_model2)
print(paste("The slope is", linear_trend_model2$coefficients[2], "and intercept is", linear_trend_model2$coefficients[1]))
linear_trend2_beta0 = as.numeric(linear_trend_model2$coefficients[1])
linear_trend2_beta1 = as.numeric(linear_trend_model2$coefficients[2])

linear_trend_model3=lm(ts_object[,4]~t)  
summary(linear_trend_model3)
print(paste("The slope is", linear_trend_model3$coefficients[2], "and intercept is", linear_trend_model3$coefficients[1]))
linear_trend3_beta0 = as.numeric(linear_trend_model3$coefficients[1])
linear_trend3_beta1 = as.numeric(linear_trend_model3$coefficients[2])
```

The slope and intercept for Total Biomass Energy production are `r linear_trend1_beta1` and `r linear_trend1_beta0` respectively.The slope and intercept for Total Renewable Energy Production are `r linear_trend2_beta1` and `r linear_trend2_beta0` respectively. The slope and intercept for Hydroelectric Power Consumption are `r linear_trend3_beta1` and `r linear_trend3_beta0` respectively.


### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?
```{r}
t = c(1: nmonths)

biomass_detrend = new_data[, 2] - (linear_trend1_beta0 + linear_trend1_beta1*t)
renewable_detrend = new_data[, 3] - (linear_trend2_beta0 + linear_trend2_beta1*t)
hydro_detrend = new_data[, 4] - (linear_trend3_beta0 + linear_trend3_beta1*t)

ggplot(data = new_data, aes(x = Date, y = new_data[, 2])) + 
             geom_line(color="green") +
             ylab(paste0(colnames(new_data)[(2)],sep="")) +
            
            geom_line(aes(y=biomass_detrend), col="blue")+
            geom_smooth(aes(y=biomass_detrend),color="red",method="lm") 

ggplot(data = new_data, aes(x = Date, y = new_data[, 3])) + 
             geom_line(color="green") +
             ylab(paste0(colnames(new_data)[(3)],sep="")) +
            
            geom_line(aes(y=biomass_detrend), col="blue")+
            geom_smooth(aes(y=biomass_detrend),color="red",method="lm") 

ggplot(data = new_data, aes(x = Date, y = new_data[, 4])) + 
             geom_line(color="green") +
             ylab(paste0(colnames(new_data)[(4)],sep="")) +
            
            geom_line(aes(y=biomass_detrend), col="blue")+
            geom_smooth(aes(y=biomass_detrend),color="red",method="lm") 
```

**Yes, all three plots had their trend lines changed. All the trend lines became horizontal meaning slope = 0. Also, from the plots, we can observe that all 3 intercepts became 0**

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?
```{r}
par(mfrow=c(1,2))  
Acf(biomass_detrend,lag.max=40,main=paste("ACF of biomass",sep=" ")) 
Pacf(biomass_detrend,lag.max=40,main=paste("PACF of biomass",sep=" "))

Acf(renewable_detrend,lag.max=40,main=paste("ACF of renewables",sep=" ")) 
Pacf(renewable_detrend,lag.max=40,main=paste("PACF of renewables",sep=" "))

Acf(hydro_detrend,lag.max=40,main=paste("ACF of hydro",sep=" ")) 
Pacf(hydro_detrend,lag.max=40,main=paste("PACF of hydro",sep=" "))
```

**With regard to ACF plot, there isn't much of a change - the ACF gradually decreases as lag increases for Total Biomass Energy Production and Total Renewable Energy Production whereas for Hydro Power Consumption, the ACF plot suggests seasonality in the data. Therefore, in addition to detrending, we'd also need to remove seasonality. Same goes with PACF plots too, marginal change, if any, on the magnitude of the correlation function. Have to explore de-seasoning to know more!**

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.
```{r}
dummies = seasonaldummy(ts_object[,2])

seas_means_model_biomass=lm(new_data[,2]~dummies)
summary(seas_means_model_biomass)
seas_means_model_renewable=lm(new_data[,3]~dummies)
summary(seas_means_model_renewable)
seas_means_model_hydro=lm(new_data[,4]~dummies)
summary(seas_means_model_hydro)

beta_int_biomass=seas_means_model_biomass$coefficients[1]
beta_coeff_biomass=seas_means_model_biomass$coefficients[2:12]
beta_int_energy=seas_means_model_renewable$coefficients[1]
beta_coeff_energy=seas_means_model_renewable$coefficients[2:12]
beta_int_hydro=seas_means_model_hydro$coefficients[1]
beta_coeff_hydro=seas_means_model_hydro$coefficients[2:12]

```

**From Seasonal means model for biomass and renewable, the aren't significant dummy variables. For hydro power consumption, there are many significant dummy variables (May, June, Sep, Oct, and Nov). This bolsters the previous argument - the prersence of seasonality in hydro power consumption data**
### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?
```{r}
biomass_seas_comp=array(0,nmonths)

for(i in 1:nmonths){
  biomass_seas_comp[i]=(beta_int_biomass+beta_coeff_biomass%*%dummies[i,])
  
}

ggplot(new_data, aes(x=Date, y=new_data[,2])) +
            geom_line(color="green") +
            ylab(paste0("Biomass")) +
            geom_line(aes(y=biomass_seas_comp), col="blue")
deseason_biomass = new_data[,2]-biomass_seas_comp

ggplot(new_data, aes(x=Date, y=new_data[,2])) +
            geom_line(color="red") +
            ylab(paste0("Biomass")) +
            geom_line(aes(y=deseason_biomass), col="orange")

energy_seas_comp=array(0,nmonths)
for(i in 1:nmonths){
  energy_seas_comp[i]=(beta_int_energy+beta_coeff_energy%*%dummies[i,])
  
}
ggplot(new_data, aes(x=Date, y=new_data[,3])) +
            geom_line(color="green") +
            ylab(paste0("Renewable Energy")) +
            geom_line(aes(y=energy_seas_comp), col="orange")
deseason_renewable_energy <- new_data[,3]-energy_seas_comp
ggplot(new_data, aes(x=Date, y=new_data[,3])) +
            geom_line(color="blue") +
            ylab(paste0("Renewable Energy")) +
            geom_line(aes(y=deseason_renewable_energy), col="red")

hydro_seas_comp=array(0,nmonths)
for(i in 1:nmonths){
  hydro_seas_comp[i]=(beta_int_hydro+beta_coeff_hydro%*%dummies[i,])
  
}
ggplot(new_data, aes(x=Date, y=new_data[,4])) +
            geom_line(color="red") +
            ylab(paste0("Hydroelectric Consumption")) +
            geom_line(aes(y=hydro_seas_comp), col="orange")
deseason_hydro = new_data[,4]-hydro_seas_comp
ggplot(new_data, aes(x=Date, y=new_data[,4])) +
            geom_line(color="green") +
            ylab(paste0("Hydroelectric Consumption")) +
            geom_line(aes(y=deseason_hydro), col="red")
```

**After de-seasoning, all 3 plots changed. The overall shape of the plots, especially for biomass and renewables, did not change much. For hydropower the de-seasoned plot is very evident in terms of increases and decreases of data. All conclusions, once again, fall in line with what we observed with Question 6**

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?
```{r}
par(mfrow=c(1,2))  
Acf(deseason_biomass,lag.max=40,main=paste("ACF of biomass",sep=" ")) 
Pacf(deseason_biomass,lag.max=40,main=paste("PACF of biomass",sep=" "))

Acf(deseason_renewable_energy,lag.max=40,main=paste("ACF of renewables",sep=" ")) 
Pacf(deseason_renewable_energy,lag.max=40,main=paste("PACF of renewables",sep=" "))

Acf(deseason_hydro,lag.max=40,main=paste("ACF of hydro",sep=" ")) 
Pacf(deseason_hydro,lag.max=40,main=paste("PACF of hydro",sep=" "))
```

**The ACF of Hydro Power Consumed, after deseasoning, shows an  obvious seasonal variation to a gradual decrease in autocorrelation with increase in lag. The PACF of Hydro Power Consumed shows a weak seasonal variation with almost zero variation after deseasoning. The ACF plots for biomass and renewables were pretty much the same but the PACF plots were showed a decrease in seasonality**

