---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 5 - Due date 02/28/22"
author: "Abbhijith Hari Gopal"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A05_Sp22.Rmd"). Submit this pdf using Sakai.

R packages needed for this assignment are listed below. Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(xlsx)
library(readxl)
library(forecast)
library(tseries)
library(ggplot2)
library(Kendall)
library(lubridate)
library(tidyverse)  #load this package so yon clean the data frame using pipes
```

## Decomposing Time Series

Consider the same data you used for A04 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review.

```{r}
#Importing data set - using xlsx package
energy_data <- read.xlsx(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow = 13,sheetIndex=1) #startRow is equivalent to skip on read.table

#Now let's extract the column names from row 11 only
read_col_names <- read.xlsx(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow = 11,endRow=11,sheetIndex=1)

colnames(energy_data) <- read_col_names
head(energy_data)

nobs=nrow(energy_data)
nvar=ncol(energy_data)

```

### Q1

For this assignment you will work only with the following columns: Solar Energy Consumption and Wind Energy Consumption. Create a data frame structure with these two time series only and the Date column. Drop the rows with *Not Available* and convert the columns to numeric. You can use filtering to eliminate the initial rows or convert to numeric and then use the drop_na() function. If you are familiar with pipes for data wrangling, try using it!

```{r}
date_of_interest = energy_data[, 1]
Solar_Wind = energy_data[, 8:9]
head(date_of_interest, 5) #As a check
head(Solar_Wind, 5) #As a check

#converting to numeric
Solar_Wind$`Solar Energy Consumption` = as.numeric(Solar_Wind$`Solar Energy Consumption`)
Solar_Wind$`Wind Energy Consumption` = as.numeric(Solar_Wind$`Wind Energy Consumption`)
class(Solar_Wind$`Solar Energy Consumption`) #As a check
class(Solar_Wind$`Wind Energy Consumption`) #As a check

#new data frame of interest
converted_df = cbind(date_of_interest, Solar_Wind)
head(converted_df, 5) #As a check
class(converted_df)#As a check

#dropping NA
Solar_Wind = drop_na(Solar_Wind)
head(Solar_Wind, 5) #As a check
```

### Q2

Plot the Solar and Wind energy consumption over time using ggplot. Plot each series on a separate graph. No need to add legend. Add informative names to the y axis using `ylab()`. Explore the function scale_x\_date() on ggplot and see if you can change the x axis to improve your plot. Hint: use *scale_x\_date(date_breaks = "5 years", date_labels = "%Y")")*

```{r}
ggplot(data = converted_df) + 
  geom_line(aes(x = converted_df[, 1], y = converted_df[, 2]), color = "Red") + 
  geom_line(aes(x = converted_df[, 1], y = converted_df[, 3]), color = "Blue") +
  ylab("Energy Consumption (Trillion Btu)") + 
  xlab("Date") + 
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") + 
  ggtitle("Solar and Wind Consumption plot")
```

### Q3

Now plot both series in the same graph, also using ggplot(). Look at lines 142-149 of the file `05_Lab_OutliersMissingData_Solution` to learn how to manually add a legend to ggplot. Make the solar energy consumption red and wind energy consumption blue. Add informative name to the y axis using `ylab("Energy Consumption)`. And use function scale_x\_date() again to improve x axis.

```{r}
ggplot(data = converted_df) + 
  geom_line(aes(x = converted_df[, 1], y = converted_df[, 2], color = "Solar")) + 
  geom_line(aes(x = converted_df[, 1], y = converted_df[, 3], color = "Wind")) +
  ylab("Energy Consumption (Trillion Btu)") + 
  xlab("Date") + 
  scale_colour_manual("", 
                      breaks = c("Solar", "Wind"),
                      values = c("Solar"="red", "Wind"="blue")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") + 
  ggtitle("Solar and Wind Consumption plot")
```

### Q3

Transform wind and solar series into a time series object and apply the decompose function on them using the additive option, i.e., `decompose(ts_data, type = "additive")`. What can you say about the trend component? What about the random component? Does the random component look random? Or does it appear to still have some seasonality on it?

```{r}
#converting to time series object
converted_df_ts = ts(converted_df, frequency = 12)
class(converted_df_ts) #As a check

#Decomposing
Solar_decompose = decompose(converted_df_ts[, 2], type = "additive")
Wind_decompose = decompose(converted_df_ts[, 3], type = "additive")

#plotting the decomposed data
plot(Solar_decompose)
plot(Wind_decompose)
```
**The trend appears to be increasing for both solar and wind consumption - positive trend. The random component has seasonality in it for both solar and wind consumption although, its more conspicuous for solar than wind. **


### Q4

Use the decompose function again but now change the type of the seasonal component from additive to multiplicative. What happened to the random component this time?

```{r}
#Decomposing
Solar_decompose = decompose(converted_df_ts[, 2], type = "multiplicative")
Wind_decompose = decompose(converted_df_ts[, 3], type = "multiplicative")

#plotting the decomposed data
plot(Solar_decompose)
plot(Wind_decompose)
```
**The random component has become a lot better in that the seasonality has been removed, although not completely, due to the introduction of type="multiplicative"for both solar and wind consumption.**


### Q5

When fitting a model to this data, do you think you need all the historical data? Think about the data from 90s and early 20s. Are there any information from those years we might need to forecast the next six months of Solar and/or Wind consumption. Explain your response.

> Answer: Well, it doesn't look like we need all of the histroical data from 90s. It looks like we can just start from 2000 for the wind power consumption and start from 2010 for the solar power consumption. the reason for this is because, prior to those dates, the trend polts for both of 'em remains constant and remain closer to 0. Therefore, any further data prior won't help in better fitting the model and forecasting future values. 

### Q6

Create a new time series object where historical data starts on January 2012. Hint: use `filter()` function so that you don't need to point to row numbers, .i.e, `filter(xxxx, year(Date) >= 2012 )`. Apply the decompose function `type=additive` to this new time series. Comment the results. Does the random component look random? Think about our discussion in class about trying to remove the seasonal component and the challenge of trend on the seasonal component.

```{r}
new_converted_df = converted_df %>%
  filter(year(converted_df[, 1]) >= 2012)

head(new_converted_df, 50) #As a check

#converting to time series object
converted_df_new_ts = ts(new_converted_df, frequency = 12)
head(converted_df_new_ts, 5) #As a check
class(converted_df_new_ts) #As a check

#Decomposing
Solar_new_decompose = decompose(converted_df_new_ts[, 2], type = "additive")
Wind_new_decompose = decompose(converted_df_new_ts[, 3], type = "additive")

#plotting the decomposed data
plot(Solar_new_decompose)
plot(Wind_new_decompose)
```

> Answer: There is a massive shift now. First. the trend, as before is indeed increasing but, this time we see a much steady increase for both solar and wind consumption. With regard to random component, there is seasonality in both the solar consumption and wind consumption but it is very apparent in solar than it is for wind. Differencing is one possible way to remove seasonlity at this point. 


